---
title: "lab_03"
author: "Derek Willis/ Rob Wells"
date: "2025-09-24"
output: html_document
---
---
student_name: Raphael Romero Ruiz
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TASK: Load libraries

```{r}
# Turn off scientific notation
options(scipen=999)

library(tidyverse)
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab that are in the data folder: the Maryland payments to vendors and Maryland grants and loans AND, using janitor, change the column names so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows
md_payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv", guess_max=10)
cleaned_md_payments <- md_payments |>
  clean_names()

# Load in the Maryland grants and loan data
grants_and_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv", guess_max=10)
cleaned_md_grants_and_loans <- grants_and_loans |>
  clean_names()

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed in the payments data. 

Look at & search the records in the dataframe to start your search, starting with a broad search and then narrowing it down. Then write code to find rows where the vendor name contains some version of the county's school system.

HINT: you should use filter() just once, and inside your filter you want to look for the variations one at a time. Think "this version OR that version."

When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find?

A1:
There are 70 records of payments to the Prince George's County public school system. 

```{r}
pg_schools <- cleaned_md_payments %>% 
  filter(str_detect(vendor_name, "PG COUNTY PUBLIC SCHOOL|PRINCE GEORGE'S COUNTY PUBLIC SCHOOLS|PRINCE GEORGES COUNTY PUBLIC SCHOOL|PRINCE GEORGES COUNTY SCHOOL|PRINCE GEORGES COUNTY PUBLIC SCHOOLS|PRINCE GEORGES PUBLIC SCHOOLS"))
```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2:
The 2021 fiscal year had the most total spending amount with $1,470,133,533. It's interesting that the top three highest amounts have happened in order, indicating that the State of Maryland has been increasing it's spending towards PG public school system every year. 
```{r}
pg_schools %>% 
  group_by(fiscal_year) %>% 
  summarise(total_amt = sum(amount, na.rm=TRUE)) %>%
  arrange(desc(total_amt))
```

Q3. This question is in two parts. 

Part 1: Code: You're interested in examining the vendors paid for capital projects at the University of Maryland. Write code that does the following: 
--Filters payments data to capital projects at the University of Maryland
--Produce a list of vendors and the total amount they received. 
Part 2: Provide a written answer, not code.

Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? 
--Provide at least three examples of vendor names that are similar and should be combined into one. 
--Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3.
In order to get a clean dataframe where there is only one record per vendor and zip code, I would need to standardize the vendor names. To do this I would rely on something like "str_to_upper" to manipulate my strings doing things removing punctuation and putting words in standard case. This way a vendors like "WHITING TURNER CONTRACTING CO" and "WHITING TURNER CONTRACTING COMPANY" or "ZULEIKA M GHODSI MD PC" and "ZULEIKA M GHODSI MD PC  DBA DELMARVA LASER EYE CTR" or "ZOOM" and "ZOOM VIDEO COMMUNICATIONS, INC." would treated as one. Once the names are uniform, I would be able to group the data by the cleaned vendor name and zip code to ensure accurate totals.

```{r}
capital_vendor_totals <- cleaned_md_payments %>% 
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) %>% 
  group_by(vendor_name) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE)) %>%
  arrange(desc(total_amount))
```
```{r}
#my attempt at doing the explanation but in code
cleaned_vendors_totals <- cleaned_md_payments %>%
  #consulted perplexity here, learned about str_remove_all,str_replace_all and str_squish
  mutate(
    vendor_clean = str_to_upper(vendor_name),
    vendor_clean = str_remove_all(vendor_clean, "[,\\.']"),
    vendor_clean = str_replace_all(vendor_clean, "\\b(INC|LLC|CO|CORP|GROUP)\\b", ""),
    vendor_clean = str_squish(vendor_clean)
  ) %>%
  group_by(vendor_clean) %>%
  summarise(total_amount = sum(amount, na.rm = TRUE))
```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. 

Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. 
Then create a total amount of money for each fiscal year. 

Write a sentence summarizing how stem cell research funding has changed over time. Be specific with the totals.

A4:
The funding for STEM cell research has fluxated over time peaking with a huge fiscal budget of 14,643,721 in 2009 and then a deep cut of more than 50% the following year. Funding increased between 2011 and 2013, but then decreased hovering between 6,254,308 and 6,785,990 from 2014 to 2018. The following two years saw a break from the $7M cap, with spending in 2019 at 7,799,987 and 2020 at 7,125,063. Then 2021 saw a dramatic drop to 4,832,225. The following year saw a return to normalcy at a familiar 7,069,212. The latest figure sits at 12,415,121 echoing the fiscal spending seen in 14 years prior to 2023. 

```{r}
cleaned_md_grants_and_loans %>% 
  mutate(description_clean = str_to_upper(description)) %>%
  filter(str_detect(description_clean, "STEM CELL")) %>% 
  group_by(fiscal_year) %>% 
  summarise(total_amount = sum(amount, na.rm=TRUE))
```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5:
I have a total of 18 rows of grantees from the 20742 area code. Some of these are duplicates, mainly related to the multiple names associated with UMD College Park, for example: UNIVERSITY OF MD,COLLEGE PARK and University of Maryland College Park. Putting together the dupes manually, I count 16 individual grantees. In order to clean the results I would have to include a casewhen() line here to group all of the various names for UMD under the grantee column. The grantee most interesting to me is UMD Data Center project, the Maryland Energy Administration gave 159,441 to develop energy efficient initiatives. With all of the discussion about how much energy is used at this types of infrastructure projects I wonder how this will be different,
```{r}
grantees_umd_cp <- cleaned_md_grants_and_loans %>% 
  filter(str_detect(zip_code, "20742")) %>% 
  group_by(grantee) %>% 
  summarise(total_amount = sum(amount, na.rm = TRUE))
```
