---
title: "lab_01"
author: "Derek Willis / Rob Wells"
date: "2025-09-10"
output: html_document
---

Raphael Romero Ruiz 9/10/2025



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to: \* run existing code as directed (look for **Task**). \* modify existing code as directed (look for **Task**). \* write code in empty codeblocks provided to answer questions included (look for **Q**). \* write out the answer in the form of a complete sentence in the space given (look for **A**). When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

Remember to add any new verbs or concepts to your class reference notebook!

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data

We're going to load the UMD courses data.

**Task** Run the codeblock below to load the data, which is contained in an RDS file.

```{r}
umd_courses <- read_rds("lab_01.rds")
```

## Answer questions

**Q1** Which department has the highest mean (average) number of seats for its classes? You need to show the department with the highest average first. Which one is the most surprising/interesting to you? 

**A1**
The Computer, Math, and Natural Sciences department has an average of 440 seats, which is interesting to me because all of those seats are actually just in a single class rather than spread across like in other departments. 
```{r}
umd_courses |>
  group_by(department) |>
  summarise(
    count_classes = n(),
    total_seats = sum(seats),
    mean_seats = mean(seats),
  ) |>
  arrange(desc(mean_seats))
```

**Q2** Find out the greatest number of times any individual class has been offered in our dataset. Your result should sort to show the greatest number of classes first. If several classes are tied for the most offered, describe how many classes are ranked in the top category for offerings.

```{r}
umd_courses |>
  group_by(id) |>
  summarise(
    count_classes = n()
  ) |>
arrange(desc(count_classes))

```


**A2** About 145 classes are tied with the most amount of classes offered across academic terms.



**Q3** Count the number of courses for each instructor and department. Which instructor has taught the most? HINT: To filter out blank and missing values from a column, drop_na() is your friend.

```{r}
umd_courses |>
  drop_na(instructors) %>% 
  group_by(instructors, department) %>% 
  summarise(
    count_classes = n()
  ) %>% 
  arrange(desc(count_classes))
```

#test
umd_courses_split <- umd_courses |>
  drop_na(instructors) |>
  mutate(
    instructors_split = str_split_fixed(instructors, ";", 6)
  ) |>
  unnest_wider(instructors_split, names_sep = "_")

**A3** The instructor that has taught the most classes is Keith Yearwood, he is the only human in the top 3. 

```{r}
```

**Q4** Given the results for Q3, what would have made it easier to find the answer? You don't have to know *how* to accomplish this; I want you to describe in words what would make the answer easier to see. How confident are you that the answer you got is the right one, and why? 
**A4**
Being able to filter out classes with a "Instructor:TBA" character value would have allowed me to see Yearwood at the very top of the table. I am confident this is the correct guy because 

**Q5** Take one of the questions you listed in the final question of the pre_lab and try to answer it here in code. If you don't know how to fully answer it, write what code you can and describe in words what else you would need to do. If you got an answer, is it interesting? 

**A5** My question was: Which department has the highest number of seats? The answer would be University Courses.

```{r}
umd_courses %>% 
  group_by(id, seats, department) %>% 
  summarise(
    count_classes = n()
  ) %>% 
arrange(desc(seats))

```
